High-Performance Language Primitives for Distributed Architectures using Asynchronous Events

### 2012-07-18 doc/asplos2013/impl.tex
Text for the event implementation section

To support inter-node operations our implementation relies on GASNet active
messages for communication.  Active messages are a command and a payload
that are sent remotely and cause a handler to be run on the target node.
A simple example of our use of active messages can be illustrated by the 
{\tt spawn} call on a {\tt Processor} object.  Consider that {\tt spawn}
is invoked on a processor that is on a remote node from the node on which
the currently running task is executing.  This operation is converted into
an active message that contains all the information about the task launch.
When the active message is handled by the remote node, the task and all its
metadata are registered with the processor on which the task was launched.
While most of the implementation of our heterogeneous runtime is straight
forward, the next three sections describe in greater detail the nuances
of the implementation of events, locks, and reduction instances.

\subsection{Event Implementation}
\label{subsec:eventimpl}

>>>>

To support inter-node operations our implementation relies on GASNet active
messages for communication.  Active messages are a command and a payload
that are sent remotely and cause a handler to be run on the target node.
A simple example of our use of active messages can be illustrated by the 
{\tt spawn} call on a {\tt Processor} object.  Consider the case where {\tt spawn}
is invoked on a processor that is on a remote node.  This operation is converted into
an active message that contains all the information about the task launch.
When the active message is handled by the remote node, the task and all its
metadata are registered with the processor on which the task was launched.
While most of the implementation of our heterogeneous runtime is straight
forward, the next three sections describe in greater detail the nuances
of the implementation of events, locks, and reduction instances.

\subsection{Event Implementation}
\label{subsec:eventimpl}
Events are created on demand by each node.  The ID of the creating node is encoded
in the upper bits of each event's ID and the creating node is said to be the 
event's {\em owner}.  If an event is used as an argument to an
operation on a remote node then the remote node can determine the owner of 
the event and send an active message to become a {\em subscriber} of the event.  When a
remote node becomes a subscriber to an event it guarantees that it will receive an active
message from the event's owner when the event triggers.  To minimize the amount
of communication that occurs between nodes, each node remembers the events to which it
is subscribed and guarantees that it only subscribes to an event once.  In the case
where there may be many remote waiters on an event 
this can dramatically reduce the number of inter-node active messages.

When an event triggers, it automatically notifies all of the operations that are waiting
on its local node.  The owner node will also send active messages to all of the subscriber nodes
telling them that the event has triggered.  Even though the event has now notified all of
its waiters, the memory on the owning node for storing the event cannot be reclaimed 
because there still may be handles to the event in the system which will request to
use this event in the future.  These requests will need to be satisfied saying that 
the event has already triggered for the duration of the application.  Rather than waste
the resources, we recycle event implementations.

A {\em dynamic event} is a logical construct that will only be triggered once and is the level
of abstraction at which the programmer reasons about events.  A {\em physical event} is an
actual event implementation that consumes resources on its owner node.  Multiple dynamic events
can be mapped to a single physical event.  However, for each physical event there can be at most
one {\em active} event in the set of dynamic events that are mapped to it.  An active event is a 
dynamic event that is yet to trigger.  By only having one active dyanmic event at a time 
there is no need to disambiguate event triggers that are sent to the physical event.  Each
physical event maintains a count of the number of trigger operations that it has observed.

When the runtime needs to return a handle corresponding to a new dynamic event, it finds
a physical event which currently has no active events (e.g. the trigger count equals the number
of dynamic events mapped to the physical event).  The runtime increases the count
of the number of dynamic events mapped to the physical event.  The handle that is returned 
corresponds to a new dynamic event which contains an ID that refers to the physical
event it is mapped to and a {\em generation} that is the dynamic event number for the particular
physical event.  Note that by definition the generation will always be one greater than the
trigger count at the time of dynamic event creation.

Within this framework it is now very easy to test whether the dynamic event specified by a
handle has triggered.  Using the ID that is contained within the handle we find the
physical event implementation (which may require an active message if the owner node is
remote from where the test is begin performed).  We can then compare the generation of
the handle to the number of observed triggers contained by the physical event.  If the
number of physical triggers is greater than or equal to the handle's generation, then
the dynamic event has triggered.  We show in Section~\ref{sec:apps} that in practice
the number of needed physical events is significantly less than the number of dynamic events.

We also leverage the mapping of dynamic events onto physical events to improve the efficiency
of subscriptions.  Nodes cache the last generation for every physical event for which they
have seen a trigger notification.  If the trigger test described above detects a trigger based
on the cached information then there would be no need for the subscription which would save an
active message.  If the event has not been detected to have
triggered locally then an active message is sent to the owner which would always have been necesary
without the dynamic to physical event mapping.
